A Stopped Negative Binomial Distribution

### 2016-07-18 stat-and-prob-letters/snb-paper.tex
Another day another draft.

If all 17 patients are enrolled at once, as in the classic
design, then the sample size is 17. However, in most clinical trials the
patients are enrolled sequentially with one patient's outcome realized
before the next one enters the trial. In the present example, observing seven
successful patients is one endpoint and the number of enrollees required
could be as small as seven. Similarly 11
observed treatment failures also ends the trial. This sampling mechanism, in
which the experiment ends as soon as any predefined endpoint is reached, is
called {\em curtailed sampling}. Under curtailed sampling the range of the
sample size for this trial is between seven and 17.

@@

\begin{figure}[bp!]
\centering
\subfloat[enrollment distribution]{\includegraphics[width=0.5\textwidth]{snb-first-plot.pdf}}
\hfill
\subfloat[mean and variance varying $p$]{\includegraphics[width=0.5\textwidth]{mean-and-variance.pdf}}
\caption{
The distribution of a trial that stops after seven patients respond to treatment
or 11 patients do not. Panel (a) shows the distribution when
$p = 0.2$. The mean and variance of the distribution varying $p$ 
between zero and one are shown in Panel (b).
}
\label{fig:exp-and-var}
\end{figure}

@@

In the rest of this work we derive the distribution of the number of 
enrollees needed
to observe either $s$ successes or $t$ failures. We refer to this distribution
as the Stopped Negative Binomial (SNB). Some of its characteristics are
summarized in Tab.~\ref{tab:snb}.
This paper derives these results
and explores properties of the distribution.
%The next section introduces our notation and basic results
%including the density of the distribution along with a description of
%its relation to other distributions. 
Section 2 derives the distribution function
based on a defined Bernoulli process and gives some basic properties.
Section 3 shows how the distribution is related to other standard
distributions and connects the SNB tail probability to the binomial tail 
probability.
Section 4 derives the moment generating function.
Section 5 derives the predictive distribution when $p$ has a beta distribution.

@@

The moment generating function for the SNB is calculated in a manner similar to 
that of two negative binomial distributions. 
\begin{prop} Let $Y$ be distributed SNB with parameters $p$, $s$, and $t$.
Then the moment generating function (MGF) of $Y$ is
\begin{equation} \label{eqn:mgf}
\mathbb{E}~e^{xY} = \left(\frac{p e^x}{1 - qe^x}\right)^s 
  \mathcal{I}_{1-qe^x} (s, t) + \left(\frac{qe^x}{1-pe^x}\right)^t 
  \mathcal{I}_{1-pe^x}(t, s)
\end{equation}
for $q = 1-p$ defined for $x \leq \min \left\{\log(1/p), \log(1/q) \right\}$.
\end{prop}
\begin{proof}
The MGF of the SNB is:
\begin{equation*}
\mathbb{E}~e^{xY} = \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s q^{k-s} e^{kx} 
  + \sum_{k=t}^{s+t-1} {k-1 \choose t-1} p^{k-t} q^t e^{kx}
\end{equation*}
and can be rewritten as:
\begin{equation} \label{eqn:first_sum}
\mathbb{E}~e^{xY} = \sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe^x)^{s} (qe^x)^{k-s} 
  + \sum_{k=t}^{s+t-1}{k-1 \choose t-1} (qe^x)^t (pe^x)^{k-t}.
\end{equation}
The first summation in (\ref{eqn:first_sum}) satisfies
\begin{align*}
\sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe)^{sx} (qe^x)^{k-s} &= 
  \left(\frac{pe^x}{1 - qe^x}\right)^s \ \ \sum_{k=s}^{s+t-1} {k-1 \choose s-1} 
    (qe^x)^{k-s} (1-qe^x)^s \\
  &= \left(\frac{pe^x}{1 - qe^x}\right)^s \mathcal{I}_{1-qe^x}(s, t).
\end{align*}
Since the incomplete beta function's subscript parameter has support on zero 
to one, we have $qe^x \leq 1$. This also shows we must restrict
$x \leq -\log(q)$.
A similar expression can be derived for the second summation in 
(\ref{eqn:first_sum}) and results in
the constraint $x \leq -\log(p)$.
\end{proof}

The SNB's ability to approximate the geometric, normal, gamma, and poisson
distributions follow from it generalizing the negative binomial distribution. 
To recover the MGF of the negative binomial consider the case where
$t \rightarrow \infty$ in (\ref{eqn:mgf}). The regularized incomplete
beta function in the first term goes to one and zero in the second term.
We are left with the MGF of the negative binomial distribution.

@@

\section{The Predictive Probability Distribution}

Let us consider the SNB distribution where $p$ has a beta distribution.
\begin{prop} \label{prop:bayesian}
The predictive distribution of the SNB
distribution when $p$ is distributed Beta($\alpha$, $\beta$) is
\begin{align} \label{eqn:posterior}
\mathbb{P} [Y = k | s, t, \alpha, \beta ] &= 
  {k-1 \choose s-1} \frac{B\left(\alpha+s, k-s+\beta \right)}{B(\alpha, \beta)} 
    \ I_{\{s \leq k \leq s+k-1\}} \ + \nonumber \\
  & {k-1 \choose t-1} 
    \frac{B\left(\alpha + k - t, t+\beta\right)}{B(\alpha, \beta)} 
    \ I_{\{t \leq k \leq s+k-1\}}.
\end{align}
\end{prop}
\begin{proof}
For notational simplicity, assume that $k$ is in the range
$\min(s,t) \leq k \leq s+t-1$. When 
this is not the case appropriate terms should be removed as dictated by the indicator functions.
\begin{align*}
f(k | s, t, \alpha, \beta) = \frac{1}{B(\alpha, \beta)} & \int_0^1 {k-1 \choose s-1} p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} + \\
 & {k-1 \choose t-1} p^{k-t+\alpha-1}\left(1-p\right)^{t+\beta-1} dp \\
= \frac{1}{B(\alpha, \beta)}  {k-1 \choose s-1} & \int_0^1  p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} dp + \\
 & \frac{1}{B(\alpha, \beta)} {k-1 \choose t-1} \int_0^1  p^{k-t+\alpha-1}\left(1-p\right)^{t+\beta-1} dp.
\end{align*}
The result in (\ref{eqn:posterior}) follows by the definition of the 
beta function.
\end{proof}

To better understand the predictive distribution, let us once again return to 
our example where $s=7$, $t=11$. However, we assume $p$ is a beta distribution
with shape parameters $\alpha = 0.2 c$ and 
$\beta = 0.8 c$ for any $c > 0$.
This parameterization allows us to examine the relationship between 
uncertainty in $p$ and the SNB.

\begin{figure}[h!]
\centering
\subfloat[mean]{\includegraphics[width=0.5\textwidth]{bayesian-sample-expectation.pdf}}
\hfill
\subfloat[variance]{\includegraphics[width=0.5\textwidth]{bayesian-sample-variance.pdf}}
\caption{
The mean and variance of the predictive SNB with $s=7$, $t=11$
with $\mathbb{E}\ p = 0.2$ varying the uncertainty 
in the distribution of $p$. A larger value of $c$
corresponds to a beta with smaller variance.
}
\label{fig:bayesian-sample-size}
\end{figure}

Fig.~\ref{fig:bayesian-sample-size} shows how the mean
and variance of the predictive distribution is related to the SNB's 
mean and variance. As $c$ increases the 
predictive distribution sample statistics converge to corresponding 
statistics for the SNB with $p = 0.2$. 
%It may be noted that the compound 
%distribution's mean
%approaches the deterministic limit slowly. This
%is because when $c$ is small, the distribution of $p$ is heavily skewed. 
%As $c$ increases this effect is minimized.
When the distribution of $p$ is centered close to zero or one the mean 
estimate approach the deterministic limit relatively slowly in the uncertainty
parameter $c$.  Because of this, we note that for estimation it may be 
preferable to select shape parameters based on the median or mode of the 
beta distribution since they are less sensitive to distributional uncertainty.

%\section{Discussion}
%
%The SNB can be used to extend clinical trial methodology
%in at least two areas. First, it can be used to construct nested criteria
%for early stopping and efficacy. Current approaches are often inspired by
%the Simon 2-stage Phase II clinical trial \cite{Simon1989} where a
%a small number of patients are enrolled for an initial stage. For patients,
%success or failure is assessed. If at least a pre-specified number of
%successes are observed then the trial enters a second stage with more
%patients. If a sufficient number of successes are not observed then the
%trial fails.

>>>>

If all 17 patients are enrolled at once, as in the classic
design, then the sample size is 17. However, in most clinical trials the
patients are enrolled sequentially over time.
In the present example, observing seven
successful patients is one endpoint and the number of enrollees required
could be as small as seven. Similarly 11
observed treatment failures also ends the trial. This sampling mechanism, in
which the experiment ends as soon as any predefined endpoint is reached, is
called {\em curtailed sampling}. Under curtailed sampling the range of the
sample size for this trial is seven to 17.

@@

\begin{figure}[bp!]
\centering
\subfloat[enrollment distribution]{\includegraphics[width=0.5\textwidth]{snb-first-plot.pdf}}
\hfill
\subfloat[mean and variance varying $p$]{\includegraphics[width=0.5\textwidth]{mean-and-variance.pdf}}
\caption{
The distribution of a trial that stops after seven patients respond to treatment
or 11 patients do not. Panel (a) shows the distribution when
$p = 0.2$. The mean and variance of the distribution varying $p$ 
between zero and one are shown in panel (b).
}
\label{fig:exp-and-var}
\end{figure}

@@

In the rest of this work we derive the distribution of the number of 
enrollees needed
to observe either $s$ successes or $t$ failures. We refer to this distribution
as the Stopped Negative Binomial (SNB). Some of its characteristics are
summarized in Tab.~\ref{tab:snb}.
This paper derives these results
and explores properties of the distribution.
%The next section introduces our notation and basic results
%including the density of the distribution along with a description of
%its relation to other distributions. 
Section 2 derives the distribution function
based on a defined Bernoulli process and gives some basic properties.
Section 3 shows how the distribution is related to other standard
distributions and connects the SNB tail probability to the binomial tail 
probability.
Section 4 derives the moment generating function.
Section 5 derives the predictive and posterior distributions when $p$ has a 
beta distribution.

@@

The moment generating function for the SNB is calculated in a manner similar to 
that of two negative binomial distributions. 
\begin{prop} Let $Y$ be distributed SNB with parameters $p$, $s$, and $t$.
Then the moment generating function (MGF) of $Y$ is
\begin{equation} \label{eqn:mgf}
\mathbb{E}~e^{xY} = \left(\frac{p e^x}{1 - qe^x}\right)^s 
  \mathcal{I}_{1-qe^x} (s, t) + \left(\frac{qe^x}{1-pe^x}\right)^t 
  \mathcal{I}_{1-pe^x}(t, s)
\end{equation}
for $q = 1-p$ defined for $x < \min \left\{\log(1/p), \log(1/q) \right\}$.
\end{prop}
\begin{proof}
The MGF of the SNB is:
\begin{equation*}
\mathbb{E}~e^{xY} = \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s q^{k-s} e^{kx} 
  + \sum_{k=t}^{s+t-1} {k-1 \choose t-1} p^{k-t} q^t e^{kx}
\end{equation*}
and can be rewritten as:
\begin{equation} \label{eqn:first_sum}
\mathbb{E}~e^{xY} = \sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe^x)^{s} (qe^x)^{k-s} 
  + \sum_{k=t}^{s+t-1}{k-1 \choose t-1} (qe^x)^t (pe^x)^{k-t}.
\end{equation}
The first summation in (\ref{eqn:first_sum}) satisfies
\begin{align*}
\sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe)^{sx} (qe^x)^{k-s} &= 
  \left(\frac{pe^x}{1 - qe^x}\right)^s \ \ \sum_{k=s}^{s+t-1} {k-1 \choose s-1} 
    (qe^x)^{k-s} (1-qe^x)^s \\
  &= \left(\frac{pe^x}{1 - qe^x}\right)^s \mathcal{I}_{1-qe^x}(s, t).
\end{align*}
Since the incomplete beta function's subscript parameter has support on zero 
to one, we have $0 \leq qe^x < 1$. This also shows we must restrict
$0 \leq x < -\log(q)$.
A similar expression can be derived for the second summation in 
(\ref{eqn:first_sum}) and results in
the constraint $0 \leq x < -\log(p)$.
\end{proof}

The SNB's ability to approximate the geometric, normal, gamma, and poisson
distributions follow from it generalizing the negative binomial distribution. 
\begin{prop}
The MGF of the SNB converges to that of the negative binomial when either
$s$ or $t$ gets large. That is
\begin{equation*}
\mathbb{E} e^{xY} \rightarrow \left( \frac{pe^x}{1-qe^x} \right)^s
\end{equation*}
as $t \rightarrow \infty$. The analogous result holds when 
$s \rightarrow \infty$.
\end{prop}
\begin{proof}
The second regularized incomplete beta term in (\ref{eqn:mgf}) can be written
in terms of a cumulative binomial distribution
\begin{equation*}
\mathcal{I}_{1-pe^x}(t, s) = \mathbb{P}\left[ B \leq s-1 \right]
\end{equation*}
where $B$ is distributed as
Binomial($t-k$, $pe^x$). By Hoeffding's inequality \cite{Hoeffding1963}
it follows that
\begin{equation} \label{eqn:hoeffding}
\mathbb{P}\left[ B \leq s-1 \right] \leq 
  \exp \left(-2\frac{((t-k)pe^x - s-1)^2}{t-k} \right).
\end{equation}
As $t$ gets large $\mathcal{I}_{1-pe^x}(t, s)$ goes to zero
Likewise, $\mathcal{I}_{1-qe^x}(s, t)$ goes
to one. The proof follows by realizing 
\begin{equation*}
0 < \frac{qe^x}{1-pe^x} < 1
\end{equation*}
over the support of $x$.
\end{proof}

@@

\section{The Posterior and Predictive Probability Distribution}

Let us consider the joint SNB distribution when $p$ has a beta distribution.
\begin{prop} \label{prop:posterior}
When $p$ is distributed Beta($\alpha$, $\beta$), the 
posterior density under the SNB model is
\begin{align*}
f(p | s, t, \alpha, \beta) &= 
  \frac{ {k-1 \choose s-1} }{B(\alpha, \beta)} p^{\alpha +s -1} 
    (1-p)^{k+\beta-s-1} + \\
  & \frac{ {k-1 \choose t-1} }{B(\alpha, \beta)} p^{k+\alpha -t -1} 
    (1-p)^{\beta+t-1}.
\end{align*}
\end{prop}
\begin{proof}
This result can be found directly by multiplying the probability mass function
in (\ref{eqn:pmf}) by the density of the Beta distribution.
\end{proof}

\begin{corollary} \label{prop:bayesian}
The predictive distribution of the SNB
distribution when $p$ is distributed Beta($\alpha$, $\beta$) is
\begin{align} \label{eqn:posterior}
\mathbb{P} [Y = k | s, t, \alpha, \beta ] &= 
  {k-1 \choose s-1} \frac{B\left(\alpha+s, k-s+\beta \right)}{B(\alpha, \beta)} 
    \ I_{\{s \leq k \leq s+k-1\}} \ + \nonumber \\
  & {k-1 \choose t-1} 
    \frac{B\left(\alpha + k - t, t+\beta\right)}{B(\alpha, \beta)} 
    \ I_{\{t \leq k \leq s+k-1\}}.
\end{align}
\end{corollary}
\begin{proof}
This result can be found by integrating $p$ over the interval zero to one and
applying the definition of the beta function.
\end{proof}
%\begin{proof}
%For notational simplicity, assume that $k$ is in the range
%$\min(s,t) \leq k \leq s+t-1$. When 
%this is not the case appropriate terms should be removed as dictated by the indicator functions.
%\begin{align*}
%f(k | s, t, \alpha, \beta) = \frac{1}{B(\alpha, \beta)} & \int_0^1 {k-1 \choose s-1} p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} + \\
% & {k-1 \choose t-1} p^{k-t+\alpha-1}\left(1-p\right)^{t+\beta-1} dp \\
%= \frac{1}{B(\alpha, \beta)}  {k-1 \choose s-1} & \int_0^1  p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} dp + \\
% & \frac{1}{B(\alpha, \beta)} {k-1 \choose t-1} \int_0^1  p^{k-t+\alpha-1}\left(1-p\right)^{t+\beta-1} dp.
%\end{align*}
%The result in (\ref{eqn:posterior}) follows by the definition of the 
%beta function.
%\end{proof}

To better understand the predictive distribution, let us once again return to 
our example where $s=7$, $t=11$. However, we assume $p$ has a beta distribution
with shape parameters $\alpha = 0.2 c$ and $\beta = 0.8 c$ for any $c > 0$.
This parameterization has mean 0.2 for all $c > 0$ and variance 
$0.16 / (c+1)$.

\begin{figure}[h!]
\centering
\subfloat[mean]{\includegraphics[width=0.5\textwidth]{bayesian-sample-expectation.pdf}}
\hfill
\subfloat[variance]{\includegraphics[width=0.5\textwidth]{bayesian-sample-variance.pdf}}
\caption{
The mean and variance of the predictive SNB with $s=7$, $t=11$
with $\mathbb{E}\ p = 0.2$ varying the uncertainty 
in the distribution of $p$. A larger value of $c$
corresponds to a beta variance.
}
\label{fig:bayesian-sample-size}
\end{figure}

As $c$ increases the 
predictive distribution sample statistics converge to corresponding 
statistics for the SNB with $p = 0.2$. 
For estimation it may be 
preferable to select shape parameters based on the median or mode of the 
beta distribution since they are less sensitive to distributional uncertainty.

%%It may be noted that the compound 
%%distribution's mean
%%approaches the deterministic limit slowly. This
%%is because when $c$ is small, the distribution of $p$ is heavily skewed. 
%%As $c$ increases this effect is minimized.
%When the distribution of $p$ is centered close to zero or one the mean 
%estimate approach the deterministic limit relatively slowly in the uncertainty
%parameter $c$.  
%\section{Discussion}
%
%The SNB can be used to extend clinical trial methodology
%in at least two areas. First, it can be used to construct nested criteria
%for early stopping and efficacy. Current approaches are often inspired by
%the Simon 2-stage Phase II clinical trial \cite{Simon1989} where a
%a small number of patients are enrolled for an initial stage. For patients,
%success or failure is assessed. If at least a pre-specified number of
%successes are observed then the trial enters a second stage with more
%patients. If a sufficient number of successes are not observed then the
%trial fails.
